The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with {\em Gaia} Data and Beyond

### 2019-05-04 paper/ms.tex
Swap two appendix sections to match order in text
Fix some abbreviations

@@

\subsection{Galactic Center}
First, one must define the center of the Galaxy. This is usually taken to be
the location of the central supermassive black hole, Sgr~A\textsuperscript{*}
\citep[e.g.][]{2004ApJ...616..872R}. From stellar motions near
Sgr~A\textsuperscript{*}, the distance from the Sun to
Sgr~A\textsuperscript{*} can be precisely measured \citep{2009ApJ...692.1075G,
2018AA...615L..15G}, with a recent measurement using near-infrared
interferometry of $8.178 \pm 0.035$~kpc \citep{2019arXiv190405721A}.

Importantly, the location of Sgr~A\textsuperscript{*} may not be equivalent to
the location of the ``dynamical Galactic center:'' the point in
three-dimensional space about which the stars in the solar neighborhood are
orbiting. This assumption, although sensible and frequently made, has not yet been
justified. It is tested in any construction of a  dynamical model where the
distance from the Sun to the Galactic center, $R_0$, is a free parameter. For
example, \citet{2015ApJ...803...80K} measured $R_0$ while modelling the
dynamics of the stream Palomar 5. Many other dynamical measurements of $R_0$
have been made (\citealt{2016ARAA..54..529B} summarize many pre-{\em Gaia}
results) but none have yet achieved a precision comparable to that of the
distance to Sgr~A\textsuperscript{*}.

@@

\begin{deluxetable*}{cccccc}
\tablecaption{Stellar and gas disk scale heights of the real Milky Way and the
simulated galaxies considered in this work.\label{tab:scale_height}}
\tablehead{\colhead{galaxy} & \colhead{cold\tablenotemark{a} gas disk} &
\colhead{stellar thin disk} & \colhead{stellar thick disk} & \colhead{cold gas
softening length} & \colhead{stellar softening length} \\
\colhead{} & \colhead{(pc)} & \colhead{(pc)} & \colhead{(pc)} & \colhead{(pc)}
& \colhead{(pc)} }
\startdata
MW\tablenotemark{b} & 40 & 300 & 900 & \nodata & \nodata \\
\mi\tablenotemark{c} & 800\tablenotemark{d} & 480 & 2000 & 12.4 & 11.2 \\
\mf\tablenotemark{c} & 360 & 440 & 1280 & 12.5 & 11.2 \\
\mm\tablenotemark{c} & 250 & 290 & 1030 &10.3 & 11.2 \\
\enddata
% \tablerefs{\citet{2016ARAA..54..529B,2018arXiv180610564S,2008ApJ...673..864J}}
\tablenotetext{a}{$T<1000$ K}
\tablenotetext{b}{\citet{2008ApJ...673..864J,2016ARAA..54..529B}}
\tablenotetext{c}{\citet{2018arXiv180610564S}}
\tablenotetext{d}{The azimuthally averaged gas vertical density profile in
\mi{} is nearly constant to this height, though individual regions show smaller
scale heights and dense clouds.}
\end{deluxetable*}

The softening lengths used in the simulations can affect the ability to
resolve the very thinnest planar structures, which in turn can affect how much
the density-based midplane varies as a function of azimuth. The Milky Way's
dense, star-forming gas disk is thought to have a scale height of about
$40\,\pc$ ($3\text{-}4$ times the cold gas softening length;
\citealt{2019ApJ...871..145A}) and the thin stellar disk a scale height of
about 300 pc (30 times the stellar softening length;
\citealt{2008ApJ...673..864J}). We therefore expect that resolution effects
may still be affecting the scale heights of these components in the
simulations, especially the cold gas. Indeed, the stellar scale heights of the
simulated galaxies are equal to or larger than the MW's while the gas scale
heights are significantly larger (although the proper basis comparison is less
clear in the case of the gas; the quoted value for the Milky Way comes from
studies of high-mass star-forming regions). As noted above, the midplanes
defined by gas and stars can be tilted with respect to one another as well,
precluding extending the precision of the gas midplane definition to the
stellar component. In this work we specifically consider azimuthal variations
in the midplane \emph{as defined by the stellar mass density}, which will
subsequently be lower in the simulations than in the MW: within 200 pc of the
Sun the measured gas and stellar volume densities in the MW are roughly
comparable at $\sim 0.04\ \Msun\
\unit{pc}{-3}$, while in the simulations they are a factor of a few lower (see
Table 3 of \citealt{2018arXiv180610564S}). We thus expect that the degree of
midplane variation we find in the simulated systems is likely to be
\emph{less} than in the real MW, since on the smallest scales the simulations
are \emph{smoother} than the real Galaxy.

Cosmological simulations of Milky Way-mass galaxies are not perfect
representations of the true Milky Way in other ways as well\footnote{The failure of cosmological simulations to exactly reproduce the Milky Way is not entirely due to limitations of the numerical model. Candidate Milky Way-like galaxies are chosen solely on their mass and isolation, for which there are a wide variety of possible galaxies with qualitatively different properties.}, as discussed in
\citet{2018arXiv180610564S}. For instance, the velocity structure of \mi{} is
closer to M31's than the Milky Way's (Loebman et al. in prep). However, in
this work we are most interested in the global properties of the potential,
and specifically in deviations from axisymmetry. From this perspective, the
simulated galaxies are actually \emph{more} axisymmetric than we might expect
of the Milky Way. While they have prominent spiral arms, none has as strong a
bar as the MW does at present day, and none has a nearby companion like the
Large Magellanic Cloud. One of the three we consider (\mf) does have an
ongoing interaction with a satellite galaxy similar to Sagittarius, which has
punched through the galactic disk outside the solar circle and induced some
warping. However, we expect that, as with the use of finite smoothing lengths,
the degree of midplane variation in the simulations is likely a factor of a
few lower than what could be expected in the MW.

@@

\subsection{Description of Milky Way-Sagittarius Interaction Simulation} 
\label{ssec:sag_sim}
In addition to the cosmological zoom-ins, we will also briefly consider results
from a toy Sagittarius encounter simulation. This simulation offers us the
ability to see how the midplane varies in a more controlled environment. The
details of the simulation are given in \citet{2018MNRAS.481..286L}, but we
briefly summarize the most relevant details here. For the Milky Way, the dark
halo is modeled as a Hernquist sphere of mass $10^{12}\,\Msun$ and scale
length of $52\,\kpc$, the disk modeled as an exponential disk with a scale
radius of $3.5\,\kpc$, scale height $0.53\,\kpc$, and mass
$6\times10^{10}\,\Msun$, a Hernquist bulge of mass $10^{10}\,\Msun$ and scale
length $0.7\,\kpc$ \citep{1990ApJ...356..359H}. Sagittarius is modelled as a
dark matter Hernquist sphere of mass \Gus{something} and scale length
\Gus{something}, along with a stellar component modeled as a Hernquist sphere
of mass $6.4\times10^8\,\Msun$ and scale radius $0.85\,\kpc$. All components
are realized with live distributions of N-body particles; the MW and Sgr are
each initialized to be in equilibrium in isolation \Gus{check!}.

@@

\appendix
\section{$\Delta R$-$\Delta x$ Relation} \label{app:deltax}
We considered the effect on actions of an inaccuracy in the distance from the
Sun to the Galactic center, which introduces an offset in the $x$ coordinate,
$\Delta x$, of each star when converting to a Galactocentric coordinate
system. In observations of nearby stars, we have that $\Delta x \sim \Delta
R$. However, for the experiment we performed in Section~\ref{ssec:quant} we
considered observations of a star throughout its entire orbit. Therefore, we
must average $\Delta R$ over the course of the orbit. We derive this relation
now.

@@

\section{Orbits} \label{app:orbits}
We plot the three orbits considered throughout the work
(Table~\ref{tab:orbits}) in Figure~\ref{fig:plot_orbits}.

>>>>

\newcommand{\sgra}{Sgr~A\textsuperscript{*}}

@@

\subsection{Galactic Center}
First, one must define the center of the Galaxy. This is usually taken to be
the location of the central supermassive black hole, Sagittarius~A\textsuperscript{*}
\citep[\sgra{}, e.g.][]{2004ApJ...616..872R}. From stellar motions near
\sgra{}, the distance from the Sun to
\sgra{} can be precisely measured \citep{2009ApJ...692.1075G,
2018AA...615L..15G}, with a recent measurement using near-infrared
interferometry of $8.178 \pm 0.035$~kpc \citep{2019arXiv190405721A}.

Importantly, the location of \sgra{} may not be equivalent to
the location of the ``dynamical Galactic center:'' the point in
three-dimensional space about which the stars in the solar neighborhood are
orbiting. This assumption, although sensible and frequently made, has not yet been
justified. It is tested in any construction of a  dynamical model where the
distance from the Sun to the Galactic center, $R_0$, is a free parameter. For
example, \citet{2015ApJ...803...80K} measured $R_0$ while modelling the
dynamics of the stream Palomar 5. Many other dynamical measurements of $R_0$
have been made (\citealt{2016ARAA..54..529B} summarize many pre-{\em Gaia}
results) but none have yet achieved a precision comparable to that of the
distance to \sgra{}.

@@

\begin{deluxetable*}{cccccc}
\tablecaption{Stellar and gas disk scale heights of the real Milky Way and the
simulated galaxies considered in this work.\label{tab:scale_height}}
\tablehead{\colhead{galaxy} & \colhead{cold\tablenotemark{a} gas disk} &
\colhead{stellar thin disk} & \colhead{stellar thick disk} & \colhead{cold gas
softening length} & \colhead{stellar softening length} \\
\colhead{} & \colhead{(pc)} & \colhead{(pc)} & \colhead{(pc)} & \colhead{(pc)}
& \colhead{(pc)} }
\startdata
Milky Way\tablenotemark{b} & 40 & 300 & 900 & \nodata & \nodata \\
\mi\tablenotemark{c} & 800\tablenotemark{d} & 480 & 2000 & 12.4 & 11.2 \\
\mf\tablenotemark{c} & 360 & 440 & 1280 & 12.5 & 11.2 \\
\mm\tablenotemark{c} & 250 & 290 & 1030 &10.3 & 11.2 \\
\enddata
% \tablerefs{\citet{2016ARAA..54..529B,2018arXiv180610564S,2008ApJ...673..864J}}
\tablenotetext{a}{$T<1000$ K}
\tablenotetext{b}{\citet{2008ApJ...673..864J,2016ARAA..54..529B}}
\tablenotetext{c}{\citet{2018arXiv180610564S}}
\tablenotetext{d}{The azimuthally averaged gas vertical density profile in
\mi{} is nearly constant to this height, though individual regions show smaller
scale heights and dense clouds.}
\end{deluxetable*}

The softening lengths used in the simulations can affect the ability to
resolve the very thinnest planar structures, which in turn can affect how much
the density-based midplane varies as a function of azimuth. The Milky Way's
dense, star-forming gas disk is thought to have a scale height of about
$40\,\pc$ ($3\text{-}4$ times the cold gas softening length;
\citealt{2019ApJ...871..145A}) and the thin stellar disk a scale height of
about 300 pc (30 times the stellar softening length;
\citealt{2008ApJ...673..864J}). We therefore expect that resolution effects
may still be affecting the scale heights of these components in the
simulations, especially the cold gas. Indeed, the stellar scale heights of the
simulated galaxies are equal to or larger than the Milky Way's while the gas scale
heights are significantly larger (although the proper basis comparison is less
clear in the case of the gas; the quoted value for the Milky Way comes from
studies of high-mass star-forming regions). As noted above, the midplanes
defined by gas and stars can be tilted with respect to one another as well,
precluding extending the precision of the gas midplane definition to the
stellar component. In this work we specifically consider azimuthal variations
in the midplane \emph{as defined by the stellar mass density}, which will
subsequently be lower in the simulations than in the Milky Way: within 200 pc of the
Sun the measured gas and stellar volume densities in the Milky Way are roughly
comparable at $\sim 0.04\ \Msun\
\unit{pc}{-3}$, while in the simulations they are a factor of a few lower (see
Table 3 of \citealt{2018arXiv180610564S}). We thus expect that the degree of
midplane variation we find in the simulated systems is likely to be
\emph{less} than in the real Milky Way, since on the smallest scales the simulations
are \emph{smoother} than the real Galaxy.

Cosmological simulations of Milky Way-mass galaxies are not perfect
representations of the true Milky Way in other ways as well\footnote{The failure of cosmological simulations to exactly reproduce the Milky Way is not entirely due to limitations of the numerical model. Candidate Milky Way-like galaxies are chosen solely on their mass and isolation, for which there are a wide variety of possible galaxies with qualitatively different properties.}, as discussed in
\citet{2018arXiv180610564S}. For instance, the velocity structure of \mi{} is
closer to M31's than the Milky Way's (Loebman et al. in prep). However, in
this work we are most interested in the global properties of the potential,
and specifically in deviations from axisymmetry. From this perspective, the
simulated galaxies are actually \emph{more} axisymmetric than we might expect
of the Milky Way. While they have prominent spiral arms, none has as strong a
bar as the Milky Way does at present day, and none has a nearby companion like the
Large Magellanic Cloud. One of the three we consider (\mf) does have an
ongoing interaction with a satellite galaxy similar to Sagittarius, which has
punched through the galactic disk outside the solar circle and induced some
warping. However, we expect that, as with the use of finite smoothing lengths,
the degree of midplane variation in the simulations is likely a factor of a
few lower than what could be expected in the Milky Way.

@@

\subsection{Description of Milky Way-Sagittarius Interaction Simulation} 
\label{ssec:sag_sim}
In addition to the cosmological zoom-ins, we will also briefly consider results
from a toy Sagittarius encounter simulation. This simulation offers us the
ability to see how the midplane varies in a more controlled environment. The
details of the simulation are given in \citet{2018MNRAS.481..286L}, but we
briefly summarize the most relevant details here. For the Milky Way, the dark
halo is modeled as a Hernquist sphere of mass $10^{12}\,\Msun$ and scale
length of $52\,\kpc$, the disk modeled as an exponential disk with a scale
radius of $3.5\,\kpc$, scale height $0.53\,\kpc$, and mass
$6\times10^{10}\,\Msun$, a Hernquist bulge of mass $10^{10}\,\Msun$ and scale
length $0.7\,\kpc$ \citep{1990ApJ...356..359H}. Sagittarius is modelled as a
dark matter Hernquist sphere of mass \Gus{something} and scale length
\Gus{something}, along with a stellar component modeled as a Hernquist sphere
of mass $6.4\times10^8\,\Msun$ and scale radius $0.85\,\kpc$. All components
are realized with live distributions of N-body particles; the Milky Way and Sagittarius are
each initialized to be in equilibrium in isolation \Gus{check!}.

@@

\appendix
\section{Orbits} \label{app:orbits}
We plot the three orbits considered throughout the work
(Table~\ref{tab:orbits}) in Figure~\ref{fig:plot_orbits}.

\section{$\Delta R$-$\Delta x$ Relation} \label{app:deltax}
We considered the effect on actions of an inaccuracy in the distance from the
Sun to the Galactic center, which introduces an offset in the $x$ coordinate,
$\Delta x$, of each star when converting to a Galactocentric coordinate
system. In observations of nearby stars, we have that $\Delta x \sim \Delta
R$. However, for the experiment we performed in Section~\ref{ssec:quant} we
considered observations of a star throughout its entire orbit. Therefore, we
must average $\Delta R$ over the course of the orbit. We derive this relation
now.

@@
